% ====== PAPER INFORMATION ======
% CONFERENCE NAME: 2026 International Conference on Software, Telecommunications and Computer Networks (SoftCOM)
% CONFERENCE PAPER: Characterizing Fairness Trade-offs in Multi-Client Adaptive Streaming Systems

% Number of Pages: 
% Regular: Not exceed 06 pages
% Format: IEEE-formatted double-column pages, including figures, tables, and references

% === Key Dates:
%   Original Paper Submission Deadline: May 20, 2026
%   Acceptance Notification: July 15, 2026
%   Conference Dates: September 22-24, 2026


\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs, multirow}
\usepackage{xcolor,colortbl}
\usepackage{soul}
\usepackage{changepage,threeparttable}


% NEW PERSONALIZED PACKAGES AND COMMANDS
% ##############################
\usepackage{booktabs} % For formal tables
\usepackage{xcolor,colortbl} % for cell colors
\usepackage{changepage,threeparttable} % for wide tables
\usepackage{float}
% \usepackage{hyperref} % For URL's references
\usepackage[bookmarks=false, hidelinks]{hyperref}
\usepackage{nohyperref}
\usepackage{subcaption}
\usepackage{minted}
\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{longtable}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}


% TEXT REVIEW COMMANDS
% ##############################
\usepackage{pifont} % http://ctan.org/pkg/pifont
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

\newcommand{\rvi}[1]{{\color{blue}{#1}}} %included text - blue highlighting
\let\vi\rvi % Alias \vi to \rvi

\newcommand{\rvII}[1]{{\color{green}{#1}}} %included text second round revision
% \newcommand{\rvII}[1]{{\color{black}{#1}}} %included text

\newcommand{\rve}[1]{{\color{red}{#1}}} % excluded text




% TICKZ FIGURES
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,calc}
\tikzset{
  >={Stealth[length=2.2mm,width=2.2mm]},
  box/.style={draw, rounded corners=2pt, thick, inner sep=6pt, align=center},
  title/.style={font=\bfseries\small},
  layer/.style={box, fill=gray!6, draw=gray!70},
  edgebox/.style={box, fill=purple!6, draw=purple!60},
  agentbox/.style={box, fill=orange!8, draw=orange!70!black},
  statebox/.style={box, fill=blue!6, draw=blue!55},
  envbox/.style={box, fill=teal!6, draw=teal!60!black},
  metricbox/.style={box, fill=teal!3, draw=teal!60!black},
  thinarrow/.style={-Stealth, very thick},
  dashedbox/.style={box, dashed, draw=gray!70, fill=gray!3},
}




% WATERMARK PERSONALIZATION
% \usepackage{draftwatermark}
% Configuração da marca d'água
% \SetWatermarkText{DRAFT} % Texto da marca d'água
% \SetWatermarkScale{1} % Tamanho da marca d'água
% \SetWatermarkColor[gray]{0.9} % Cor da marca d'água (0 = preto, 1 = branco)


\begin{document}

\title{Characterizing Fairness Trade-offs in Multi-Client Adaptive Streaming \\
% {\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
% should not be used}
% \thanks{Identify applicable funding agency here. If none, delete this.}
}

\maketitle

\begin{abstract}
\rvi{Ensuring fairness among users competing for shared network resources remains a critical challenge in multi-client adaptive video streaming. While modern adaptation strategies optimize individual Quality of Experience (QoE), they often overlook the complex trade-offs between different fairness dimensions and playback stability. This study presents an analytical characterization of these trade-offs by evaluating six diverse adaptation strategies—ranging from heuristic-based baselines to fairness-aware reinforcement learning—under identical, highly variable network conditions. We employ a multi-metric framework that jointly analyzes four complementary fairness indicators: Time-Fair QoE (TF-QoE), QoE Fairness Score (QFS), Bitrate Fairness Index (BFI), and Stability-Aware Fairness (SAF). By analyzing experimental data from a 10-client edge-assisted testbed replaying real-world 4G/3G traces, we reveal structural conflicts between quality equity and adaptation instability. Our findings demonstrate that gains in bitrate fairness often come at the expense of increased quality switches for specific users, a trade-off that is significantly amplified under constrained network conditions. This work provides a rigorous observational baseline for the design of equitable streaming systems, highlighting that effective fairness evaluation must account for the interplay between resource distribution and temporal stability.}
\end{abstract}

\begin{IEEEkeywords}
Adaptive Bitrate Streaming, Fairness Trade-offs, Multi-Client Video Streaming, QoE Evaluation, Network Variability
\end{IEEEkeywords}

\section{Introduction} \label{sec: Introduction}
    \rvi{Adaptive Bitrate (ABR) streaming dominates modern Internet traffic, yet the trade-offs between Quality of Experience (QoE) and fairness in multi-user environments remain poorly understood. While modern ABR heuristics and reinforcement learning (RL) models improve individual playback performance, they frequently overlook the multi-dimensional nature of equity when clients share constrained edge resources. 

    Most existing studies quantify fairness through single-valued indicators, such as Jain's Fairness Index, which obscure temporal disparities and stability variations. This observational study addresses this gap by systematically characterizing fairness trade-offs across four distinct dimensions: temporal quality distribution (TF-QoE), distributional fairness (QFS), bitrate allocation (BFI), and adaptation stability (SAF). By analyzing a diverse set of ABR strategies—ranging from throughput-based heuristics to fairness-aware RL—under high-variability network traces, we expose structural conflicts between different fairness metrics and their impact on system stability.

    Our analysis moves beyond post-hoc measurement to provide an analytical framework for interpreting how network conditions and adaptation logic interact. We identify specific scenarios where gains in quality fairness directly compromise playback stability, revealing that a "fair" system is not necessarily a stable or optimal one. These insights are critical for the design of next-generation edge-assisted streaming systems that must balance heterogeneous user needs.

    The main contributions of this work are: (i) a multi-metric analytical characterization of fairness-stability trade-offs in multi-client adaptive streaming; (ii) the identification of structural conflicts between bitrate equity and adaptation stability; and (iii) an evaluation of how network variability amplifies fairness disparities obscured under stable bandwidth. 
    
    The remainder of this paper is organized as follows. Section \ref{sec:Background} provides the conceptual grounding. Section \ref{sec:Related Works} reviews the literature. Section \ref{sec:tradeoff_methodology} describes the evaluation methodology. Section \ref{sec:results_discussion} presents the trade-off analysis. Finally, Section \ref{sec:Conclusion and Future Work} concludes the paper.}

\section{Background \label{sec:Background}} 
    Adaptive Bitrate (ABR) streaming enables video clients to dynamically select video representations according to observed network conditions and playback state. By adapting quality over time, ABR mechanisms aim to mitigate rebuffering events, reduce startup delays, and limit disruptive quality oscillations under fluctuating bandwidth conditions. DASH-based implementations, such as DASH.js, are widely adopted in both academic research and industrial deployments, serving as representative baselines for adaptive streaming behavior. Classical ABR strategies rely on heuristic decision rules derived from throughput estimation or buffer occupancy control~\cite{jiang2012improving,yin2015control,spiteri2020bola}. While such approaches can achieve satisfactory performance in single-client scenarios, their behavior becomes significantly more complex in multi-client environments where users compete for shared and constrained network resources. Independent adaptation decisions may lead to uneven quality allocation, increased playback instability, or prolonged low-quality exposure for some users, even when overall network utilization remains efficient. Learning-based approaches, particularly those based on reinforcement learning (RL), have been explored as alternatives to heuristic ABR control. Systems such as Pensieve \cite{mao2017neural} demonstrate that data-driven policies can capture complex adaptation dynamics and improve individual Quality of Experience (QoE). However, these approaches remain primarily QoE-centric and typically evaluate fairness only as a secondary or post-hoc metric. Moreover, commonly adopted indicators such as Jain’s Fairness Index (JFI) \cite{jain1984quantitative} provide a limited view of fairness, as they do not account for temporal dynamics, instability effects, or prolonged exposure to degraded quality. Recent studies have emphasized that fairness in adaptive streaming cannot be adequately characterized by a single metric. Complementary fairness indicators have been proposed to capture different dimensions of user experience, including distributional equity, temporal exposure, throughput allocation, and playback stability~\cite{hossfeld2017definition,hossfeld2018new,akhtar2020fairness}. Taken together, these metrics reveal that fairness is inherently multi-dimensional and may conflict with traditional QoE objectives, motivating a broader evaluation perspective when analyzing multi-client adaptive streaming systems under heterogeneous network conditions.

    
\section{Related Works \label{sec:Related Works}}
    \vi{Previous works by the authors have explored QoE optimization and fairness evaluation in adaptive streaming systems~\cite{de2025joint,de2025fairness}, providing the experimental basis upon which the present analytical study is conducted. Early work on Adaptive Bitrate (ABR) streaming focused on heuristic strategies based on throughput estimation and buffer occupancy control. Representative approaches such as FESTIVE~\cite{jiang2012improving} and BOLA~\cite{spiteri2020bola} improved playback stability and efficiency under variable network conditions. However, these methods rely on independent client-side decisions and do not explicitly address fairness among multiple users competing for shared bandwidth. Reinforcement learning (RL) has been explored as an alternative paradigm, such as Pensieve~\cite{mao2017neural}, aiming to capture complex adaptation dynamics. While these approaches improve individual QoE, fairness is typically considered only as an evaluation metric. In parallel, several studies have introduced fairness-aware mechanisms. Metrics such as the QoE Fairness Score (QFS)~\cite{hossfeld2018new} and the Bitrate Fairness Index (BFI)~\cite{li2023achieving} propose alternative fairness formulations. Recent developments in 2024 and 2025 explicitly examine fairness-aware resource allocation in edge and MEC environments~\cite{xiao2024qoe,li2023edge-fair}. Despite these advances, prior work largely focuses on improving specific performance objectives or incorporating fairness as a single aggregated criterion. 

    The selection of ABR algorithms in this study (DASH-Default, BOLA, Pensieve) covers the three dominant paradigms in adaptive streaming: heuristic throughput-based, state-of-the-art buffer-based, and data-driven reinforcement learning. This diverse selection allows characterizing how fundamentally different decision logics interact with multi-dimensional fairness metrics under identical network constraints. Table~\ref{tab:comparison_related_works} summarizes representative studies, highlighting the shift toward multi-metric evaluation that motivates this work.}

    \begin{table*}[t]
        \centering
        \caption{Comparison of Related Works from a Fairness Evaluation Perspective}
        \label{tab:comparison_related_works}
        \vi{\begin{tabular}{|p{0.6cm}|p{2.9cm}|p{3.2cm}|p{4.6cm}|p{4.6cm}|}
        \hline
        \textbf{Work} & \textbf{Adaptation Strategy} & \textbf{Fairness Perspective} & \textbf{Evaluation Limitations} & \textbf{Evaluation Scope} \\
        \hline
        \cite{jiang2012improving} & Heuristic ABR & Implicit fairness via bandwidth & No explicit fairness metrics & Early analysis of fairness side-effects \\
        \cite{spiteri2020bola} & Heuristic ABR & QoE stability-focused & Fairness not evaluated & Robust buffer-based adaptation \\
        \cite{mao2017neural} & RL-based ABR & QoE-centric optimization & Fairness not considered in decision & Learned bitrate for individual QoE \\
        \cite{hossfeld2018new} & Metric proposal & QoE distribution fairness (QFS) & Metric-level evaluation only & Definition of QoE fairness metric \\
        \cite{li2023achieving} & Metric proposal & Bitrate fairness (BFI) & Single-metric evaluation & Bitrate-based fairness assessment \\
        \cite{yuan2022muabr} & Multi-agent RL & Fairness-aware coordination & Limited fairness metrics & Coordinated decisions in multi-user \\
        \cite{xiao2024qoe} & MEC optimization & QoE fairness (JFI) & Analytical modeling focus & Resource allocation in MEC (2024) \\
        \cite{li2023edge-fair} & Edge-assisted ABR & Fairness via JFI & No trade-off analysis & Fairness-constrained ABR at edge \\
        \hline\hline
        \textbf{This Work} & \textbf{Analytical Study} & \textbf{Multi-metric characterization} & \textbf{Observational focus} & \textbf{Characterizing fairness trade-offs} \\
        \hline
        \end{tabular}}
    \end{table*}


\section{Trade-off Analysis Methodology}
\label{sec:tradeoff_methodology}
    \vi{This study employs a multi-metric observational approach to characterize fairness trade-offs. The methodology focuses on the interaction between QoE indicators and four complementary fairness dimensions, using data collected from a controlled edge-assisted testbed.}

    \subsection{Experimental Setup and Scenario Definition}
        \vi{The analysis is based on experiments involving 10 concurrent DASH clients executed in Docker containers. These clients share a constrained link emulated via Mahimahi, replaying real-world cellular traces: (i) \textit{Verizon-LTE}, representing high-capacity but variable throughput ($\mu \approx 20$ Mbps), and (ii) \textit{UMTS}, capturing bandwidth-limited conditions ($\mu \approx 2$ Mbps). Table~\ref{tab:exp03-summary} summarizes the experimental parameters.}

        \begin{table}[h]
            \centering
            \caption{Summary of Experimental evaluation scenario.}
            \label{tab:exp03-summary}
            \scriptsize
            \vi{\begin{tabular}{p{2.6cm}|p{4.2cm}}
            \hline
            \textbf{Aspect} & \textbf{Description} \\
            \hline
            Scenario Type & Multi-client adaptive streaming under shared edge \\
            \hline
            Clients & 10 concurrent DASH clients in Docker containers \\
            \hline
            Streaming Engine & DASH.js player automated via Selenium \\
            \hline
            Network Traces & Verizon-LTE ($\mu \approx 20$ Mbps) and UMTS ($\mu \approx 2$ Mbps) \\
            \hline
            Video Content & Big Buck Bunny (6 levels: 360p--2160p) \\
            \hline
            Session & 120 seconds of playback per client \\
            \hline
            \end{tabular}}
        \end{table}

    \subsection{Multi-Metric Fairness Characterization}
        \vi{To move beyond single-index evaluation, we define four fairness indicators:
        \begin{itemize}
            \item \textbf{TF-QoE (Time-Fair QoE):} Quantifies temporal deviation from the session average QoE ($\bar{Q}$):
            \begin{equation}
                TF\text{-}QoE_i = 1 - \frac{|Q_i - \bar{Q}|}{\bar{Q} + \epsilon}
            \label{eq:tf_qoe}
            \end{equation}
            \item \textbf{QFS (QoE Fairness Score):} JFI applied to temporal average QoE ($\hat{Q}_i$):
            \begin{equation}
                QFS = \frac{\left(\sum_{i=1}^{N} \hat{Q}_i\right)^2}{N \cdot \sum_{i=1}^{N} \hat{Q}_i^2}
            \label{eq:qfs}
            \end{equation}
            \item \textbf{BFI (Bitrate Fairness Index):} JFI applied to average numeric representation levels ($\bar{L}_i$):
            \begin{equation}
                BFI = \frac{\left(\sum_{i=1}^{N} \bar{L}_i\right)^2}{N \cdot \sum_{i=1}^{N} \bar{L}_i^2}
            \label{eq:bfi}
            \end{equation}
            \item \textbf{SAF (Stability-Aware Fairness):} Measures equity in adaptation stability using quality switch counts ($S_i$):
            \begin{equation}
                SAF = 1 - \frac{|S_i - \bar{S}|}{\bar{S} + \epsilon}
            \label{eq:saf}
            \end{equation}
        \end{itemize}}

    \subsection{Considered ABR Strategies}
        \vi{The analysis compares three heuristic baselines (DASH-Default, BOLA, DASH-QoE) and two RL-based strategies (Pensieve Baseline and the fairness-aware variant Pensieve-Fair). This selection covers the breadth of adaptation logics used in modern streaming.}

\section{Results and Discussion}
\label{sec:results_discussion}
        \vi{The results reveal consistent behavioral patterns that differentiate adaptation strategies across fairness and stability dimensions.}

    \subsection{Fairness Metric Behavior}
        \vi{Figure~\ref{fig:fairness_metrics_exp3} illustrates that while QFS remains high ($\approx 0.9$) for most strategies, substantial disparities emerge in TF-QoE and BFI. Strategies maximizing individual QoE often suffer from poor BFI ($0.7$–$0.8$), indicating that equitable perceived quality does not necessarily imply fair resource allocation.}

    \subsection{Multi-Metric Interaction}
        \vi{Spearman correlation (Fig.~\ref{fig:fairness_multimetric_interaction_exp3}) confirms structural conflicts. TF-QoE and QFS are highly correlated ($\rho \approx 0.94$), but both conflict with BFI ($\rho \approx -0.77$). Notably, SAF is orthogonal to other metrics, showing that adaptation stability is a distinct fairness dimension.}

    \subsection{Fairness-Stability Trade-offs}
        \vi{Figure~\ref{fig:fairness_stability_exp3} demonstrates the conflict between fairness and stability. DASH-Default achieves moderate TF-QoE ($\approx 0.71$) but poor stability (SAF $\approx 0.68$). BOLA, conversely, balances both (TF-QoE $\approx 0.77$, SAF $\approx 0.79$). This trade-off is structural and persists across network conditions.}

    \subsection{Impact of Network Variability}
        \vi{Network dynamics (Fig.~\ref{fig:network_variability_exp3}) act as an amplifier for fairness trade-offs. In high-variability traces (Verizon-LTE), aggressive ABR strategies exhibit standard deviation spikes exceeding 10~Mbps, which prolongs periods of degraded quality for subsets of users. This temporal unfairness is often hidden in session-averaged metrics, underscoring the value of the SAF and TF-QoE indicators.}

\section{Conclusion \label{sec:Conclusion}}
    \vi{Fairness in multi-client adaptive streaming is inherently multi-dimensional. This observational study identifies structural trade-offs between quality equity and playback stability, showing that no single strategy dominates all metrics. Future work will explore multi-objective RL policies informed by these multi-metric insights.}

\bibliography{references}
\bibliographystyle{IEEEtran}
\end{document}
